---
title: "IHSD 7440 Assignment 3"
author: "Michaela Proffitt"
date: "2023-03-24"
output: 
  html_notebook: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction


```{r preamble, warning= FALSE}
#----------------------------------------------------------------------------#
#                                                                            #
# Assignment 3 script     #
#                                                                            #
#   Michaela Proffitt March 24th 2023  mproffitt@tulane.edu                            #
#                                                                            #
#    This script is a demonstration of the basic principles of               #
#    developing a literate program as well as some basics of conducting      #
#    descriptive statistics on select DHS data using correct sampling        #
#    weights and and analytical approaches                                   #
#                                                                            #
#                                                                            #
#----------------------------------------------------------------------------#

# loading required libraries ----
library(tidyverse)
library(haven)
library(formatR)
library(Hmisc)
#install.packages("survey") 
library(survey)
#install.packages("srvyr")
#library(srvyr)
#install.packages("weights")
library(weights)
#set any other options ----

```

## Compile date and session info

```{r docinfo, echo = FALSE, warning= FALSE}

# print and get runtime ----
date()

# session info ----
sessionInfo()

```
## Step 1: Import and investigate both datasets

```{r load, tidy = TRUE}



zam_child <- read.csv(file = "C:\\Users\\Mikki\\OneDrive - Tulane University\\Documents\\sampling\\GitHub\\IHSD_7440_HH_Sampling\\2007_Zambia_Child_2023.csv")
zam_hh <- read.csv(file = "C:\\Users\\Mikki\\OneDrive - Tulane University\\Documents\\sampling\\GitHub\\IHSD_7440_HH_Sampling\\2007_Zambia_HH_2023.csv")

# View data set in new window'
View(zam_child)
View(zam_hh)

```


```{r elements2}
describe(zam_hh)
```
```{r elements}
describe(zam_child)
```
Now to rename variables

```{r mames1}
names(zam_hh)
names(zam_hh)[names(zam_hh) == "HV022"] <- "Sample_stratum_number"
names(zam_hh)[names(zam_hh) == "HV023"] <- "Sample_domain"
names(zam_hh)[names(zam_hh) == "HV002"] <- "Household_number"
names(zam_hh)[names(zam_hh) == "HV005"] <- "Sample_weight"
names(zam_hh)[names(zam_hh) == "HV009"] <- "Number_HH_members"
names(zam_hh)[names(zam_hh) == "HV014"] <- "Number_children_5_under"
names(zam_hh)[names(zam_hh) == "HV021"] <- "Primary_sampling_unit"
names(zam_hh)[names(zam_hh) == "HV206"] <- "Has_electricity"
names(zam_hh)[names(zam_hh) == "HML1"]  <- "Number_ITNs"
names(zam_hh)[names(zam_hh) == "HV237"] <- "Water_treated"
names(zam_hh)[names(zam_hh) == "SH125A"] <- "IRS"
names(zam_hh)[names(zam_hh) == "HV237B"] <- "Water_treated_blch_chlr"
names(zam_hh)

describe(zam_hh)
```
```{r mames}
names(zam_child)
names(zam_child)[names(zam_child) == "V022"] <- "Sample_stratum_number"
names(zam_child)[names(zam_child) == "V023"] <- "Sample_domain"
names(zam_child)[names(zam_child) == "V002"] <- "Household_number"
names(zam_child)[names(zam_child) == "V005"] <- "Sample_weight"
names(zam_child)[names(zam_child) == "V009"] <- "Number_HH_members"
names(zam_child)[names(zam_child) == "V014"] <- "Number_children_5_under"
names(zam_child)[names(zam_child) == "V021"] <- "Primary_sampling_unit"
names(zam_child)[names(zam_child) == "V206"] <- "Has_electricity"
names(zam_child)[names(zam_child) == "ML1"]  <- "Number_ITNs"
names(zam_child)[names(zam_child) == "V237"] <- "Water_treated"
names(zam_child)[names(zam_child) == "H125A"] <- "IRS"
names(zam_child)[names(zam_child) == "V237B"] <- "Water_treated_blch_chlr"
names(zam_child)

```


```{r clusters2}
length(unique(zam_hh$Primary_sampling_unit)) 
# this line of code will return the number of unique observations (i.e number of clusters) within your primary sampling unit variable (remember clusters are our PSU here)
# ```
```

```{r clusters3}
length(unique(zam_child$Primary_sampling_unit)) 

```

3.) How many survey domains are there in this data set? 

```{r domains2}
length(unique(zam_hh$Sample_domain)) 

range=as.data.frame(table(zam_hh$Primary_sampling_unit))
min(range$Freq)
max(range$Freq)
```

## Step 2: Calculate proportions and standard errors for select indicators assuming SRS without weights


```{r set weight}

zam_hh$pw<- zam_hh$Sample_weight/1000000 #Here we are creating a new variable "pw" in our primary data set that corresponds with the calculated weight

zam_child$pw<- zam_child$Sample_weight/1000000 #Here we are creating a new variable "pw" in our primary data set that corresponds with the calculated weight

```



```{r SRS res, warning = FALSE, message = FALSE}

Urban.ITN <- subset(zam_hh, subset=(HH_residence=="Urban"))#creates subset featuring only HHs in urban areas
attach(Urban.ITN)

Rural.ITN <- subset(zam_hh, subset=(HH_residence=="Rural")) #creates subset featuring only HHs in rural areas
attach(Rural.ITN)# make objects in data frames accessible (rural subset) without actually typing the name of the data frame

#Now we want to calculate the proportion (and standard error) of HHs in urban areas that have ITNs:

mean.urban <- mean(Urban.ITN$HH_has_ITN)
mean.urban 
se.urban <-sqrt(var(Urban.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(Urban.ITN$HH_has_ITN)))
se.urban

mean.rural <- mean(Rural.ITN$HH_has_ITN)
mean.rural 
se.rural <-sqrt(var(Rural.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(Rural.ITN$HH_has_ITN)))
se.rural
```

``` {r SRS edu, warning = FALSE, message = FALSE}
#2) Education level

#Using subset approach (means and standard errors) for each category/value under the HH_edu variable
Noedu.ITN <- subset(zam_hh, subset=(HH_head_education=="No education"))
attach(Noedu.ITN)
Primary.ITN <- subset(zam_hh, subset=(HH_head_education=="Primary"))
attach(Primary.ITN)
Secondary.ITN <- subset(zam_hh, subset=(HH_head_education=="Secondary"))
attach(Secondary.ITN)
Higher.ITN <- subset(zam_hh, subset=(HH_head_education=="Higher than"))
attach(Higher.ITN)

mean.none <- mean(Noedu.ITN$HH_has_ITN) #No education
mean.none
se.none <-sqrt(var(Noedu.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(Noedu.ITN$HH_has_ITN)))
se.none
mean.primary<- mean(Primary.ITN$HH_has_ITN)#Primary
mean.primary
se.primary <-sqrt(var(Primary.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(Primary.ITN$HH_has_ITN)))
se.primary
mean.secondary<- mean(Secondary.ITN$HH_has_ITN)#Secondary
mean.secondary
se.secondary <-sqrt(var(Secondary.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(Secondary.ITN$HH_has_ITN)))
se.secondary
mean.higher <- mean(Higher.ITN$HH_has_ITN)#Higher than
mean.higher
se.higher <-sqrt(var(Higher.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(Higher.ITN$HH_has_ITN)))
se.higher
```

``` {r SRS edu2, warning = FALSE, message = FALSE}
#2) SES

poorest.ITN <- subset(zam_hh, subset=(HH_wealth_index=="Lowest"))
attach(poorest.ITN)
poorer.ITN <- subset(zam_hh, subset=(HH_wealth_index=="Second"))
attach(poorer.ITN)
middle.ITN <- subset(zam_hh, subset=(HH_wealth_index=="Middle"))
attach(middle.ITN)
richer.ITN <- subset(zam_hh, subset=(HH_wealth_index=="Fourth"))
attach(richer.ITN)
richest.ITN <- subset(zam_hh, subset=(HH_wealth_index=="Highes"))
attach(richest.ITN)

mean.poorest <- mean(poorest.ITN$HH_has_ITN) #No education
mean.none
se.poorest <-sqrt(var(poorest.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(poorest.ITN$HH_has_ITN)))
se.poorest
mean.poorer<- mean(poorer.ITN$HH_has_ITN)#Primary
mean.poorer
se.poorer <-sqrt(var(poorer.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(poorer.ITN$HH_has_ITN)))
se.poorer
mean.middle<- mean(middle.ITN$HH_has_ITN)#Secondary
mean.middle
se.middle <-sqrt(var(middle.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(middle.ITN$HH_has_ITN)))
se.middle
mean.richer <- mean(richer.ITN$HH_has_ITN)#Higher than
mean.richer
se.richer <-sqrt(var(richer.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(richer.ITN$HH_has_ITN)))
se.richer
mean.richest <- mean(richest.ITN$HH_has_ITN)#Higher than
mean.richest
se.richest <-sqrt(var(richest.ITN$HH_has_ITN,na.rm=TRUE)/length(na.omit(richest.ITN$HH_has_ITN)))
se.richest

mean(zam_hh$HH_has_ITN)
sqrt(var(zam_hh$HH_has_ITN,na.rm=TRUE)/length(na.omit(zam_hh$HH_has_ITN)))
```
```{r SRS res3, warning = FALSE, message = FALSE}

Urbanchild.ITN <- subset(zam_child, subset=(HH_residence=="Urban"))
attach(Urbanchild.ITN)

Ruralchild.ITN <- subset(zam_child, subset=(HH_residence=="Rural")) 
attach(Ruralchild.ITN)

#Now we want to calculate the proportion (and standard error) of HHs in urban areas that have ITNs:

mean.urbanchild <- mean(Urbanchild.ITN$Slept_ITN)
mean.urbanchild 
se.urbanchild <-sqrt(var(Urbanchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(Urbanchild.ITN$Slept_ITN)))
se.urbanchild

#And in rural areas
mean.ruralchild <- mean(Ruralchild.ITN$Slept_ITN)
mean.ruralchild #prints results
se.ruralchild <-sqrt(var(Ruralchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(Ruralchild.ITN$Slept_ITN)))#calculates standard error
se.ruralchild# prints results
```
```{r SRS edu3, warning = FALSE, message = FALSE}
poorestchild.ITN <- subset(zam_child, subset=(HH_wealth_index=="Lowest"))
attach(poorestchild.ITN)
poorerchild.ITN <- subset(zam_child, subset=(HH_wealth_index=="Second"))
attach(poorerchild.ITN)
middlechild.ITN <- subset(zam_child, subset=(HH_wealth_index=="Middle"))
attach(middlechild.ITN)
richerchild.ITN <- subset(zam_child, subset=(HH_wealth_index=="Fourth"))
attach(richerchild.ITN)
richestchild.ITN <- subset(zam_child, subset=(HH_wealth_index=="Highes"))
attach(richestchild.ITN)

mean.poorestchild <- mean(poorestchild.ITN$Slept_ITN) 
mean.poorestchild
se.poorestchild <-sqrt(var(poorestchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(poorestchild.ITN$Slept_ITN)))
se.poorestchild
mean.poorerchild<- mean(poorerchild.ITN$Slept_ITN)
mean.poorerchild
se.poorerchild <-sqrt(var(poorerchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(poorerchild.ITN$Slept_ITN)))
se.poorerchild
mean.middlechild<- mean(middlechild.ITN$Slept_ITN)
mean.middlechild
se.middlechild <-sqrt(var(middlechild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(middlechild.ITN$Slept_ITN)))
se.middlechild
mean.richerchild <- mean(richerchild.ITN$Slept_ITN)
mean.richerchild
se.richerchild <-sqrt(var(richerchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(richerchild.ITN$Slept_ITN)))
se.richerchild
mean.richestchild <- mean(richestchild.ITN$Slept_ITN)
mean.richestchild
se.richestchild <-sqrt(var(richestchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(richestchild.ITN$Slept_ITN)))
se.richestchild

mean(zam_child$Slept_ITN)
sqrt(var(zam_child$Slept_ITN,na.rm=TRUE)/length(na.omit(zam_child$Slept_ITN)))
```
``` {r SRS edu4, warning = FALSE, message = FALSE}
#2) Education level


#Using subset approach (means and standard errors) for each category/value under the HH_edu variable
Noeduchild.ITN <- subset(zam_child, subset=(Mother_education=="No education"))
attach(Noeduchild.ITN)
Primarychild.ITN <- subset(zam_child, subset=(Mother_education=="Primary"))
attach(Primarychild.ITN)
Secondarychild.ITN <- subset(zam_child, subset=(Mother_education=="Secondary"))
attach(Secondarychild.ITN)
Higherchild.ITN <- subset(zam_child, subset=(Mother_education=="Higher than"))
attach(Higherchild.ITN)

mean.nonechild <- mean(Noeduchild.ITN$Slept_ITN) #No education
mean.nonechild
se.nonechild <-sqrt(var(Noeduchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(Noeduchild.ITN$Slept_ITN)))
se.nonechild
mean.primarychild<- mean(Primarychild.ITN$Slept_ITN)#Primary
mean.primarychild
se.primarychild <-sqrt(var(Primarychild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(Primarychild.ITN$Slept_ITN)))
se.primarychild
mean.secondarychild<- mean(Secondarychild.ITN$Slept_ITN)#Secondary
mean.secondarychild
se.secondarychild <-sqrt(var(Secondarychild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(Secondarychild.ITN$Slept_ITN)))
se.secondarychild
mean.higherchild <- mean(Higherchild.ITN$Slept_ITN)#Higher than
mean.higherchild
se.higherchild <-sqrt(var(Higherchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(Higherchild.ITN$Slept_ITN)))
se.higherchild


```
```{r SRS edu5, warning = FALSE, message = FALSE}
zerochild.ITN <- subset(zam_child, subset=(Child_age=="0"))
attach(zerochild.ITN)
onechild.ITN <- subset(zam_child, subset=(Child_age=="1"))
attach(onechild.ITN)
twochild.ITN <- subset(zam_child, subset=(Child_age=="2"))
attach(twochild.ITN)
threechild.ITN <- subset(zam_child, subset=(Child_age=="3"))
attach(threechild.ITN)
fourchild.ITN <- subset(zam_child, subset=(Child_age=="4"))
attach(fourchild.ITN)

mean.zerochild <- mean(zerochild.ITN$Slept_ITN) 
mean.zerochild
se.zerochild <-sqrt(var(zerochild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(zerochild.ITN$Slept_ITN)))
se.zerochild
mean.onechild<- mean(onechild.ITN$Slept_ITN)
mean.onechild
se.onechild <-sqrt(var(onechild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(onechild.ITN$Slept_ITN)))
se.onechild
mean.twochild<- mean(twochild.ITN$Slept_ITN)
mean.twochild
se.twochild <-sqrt(var(twochild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(twochild.ITN$Slept_ITN)))
se.twochild
mean.threechild <- mean(threechild.ITN$Slept_ITN)
mean.threechild
se.threechild <-sqrt(var(threechild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(threechild.ITN$Slept_ITN)))
se.threechild
mean.fourchild <- mean(fourchild.ITN$Slept_ITN)
mean.fourchild
se.fourchild <-sqrt(var(fourchild.ITN$Slept_ITN,na.rm=TRUE)/length(na.omit(fourchild.ITN$Slept_ITN)))
se.fourchild
```
## Step 3: Calculate proportions and standard errors of select indicators assuming SRS with weights

We now want to calculate weighted  proportions assuming an SRS sampling strategy. we want to apply weights when calculating our standard errors. 

To calculate proportions AND standard errors assuming SRS with weights, we can create a function where we tell R Studio how we want our proportions to be calculated given our sampling strategy.

``` {r SRS weight}

DHS_SRS_wtdhh<-svydesign(ids= ~1, weights=~pw,data=zam_hh, nest=TRUE) #inclusion of ids= ~1 indicates no clustering (remember we are still assuming SRS at this point)

total.srswt.hh<-svymean(~HH_has_ITN, DHS_SRS_wtdhh)
total.srswt.hh
```

We can then calculate proportions and weighted standard errors, given some criteria (Residence or education level,for example), in the same table (i.e. we don't have to individually calcluate each proportion and standard error as we did in Step 1). The "svyby" command is part of the svyr package, which was loaded at the beginning of the .rmd file. 

``` {r SRS wtd prop}

reshh.SRS <- svyby(~HH_has_ITN, ~HH_residence, DHS_SRS_wtdhh, svymean, na.rm=TRUE) #Residence
reshh.SRS

seshh.SRS<- svyby(~HH_has_ITN, ~HH_wealth_index, DHS_SRS_wtdhh, svymean, na.rm=TRUE)
seshh.SRS

eduhh.SRS<- svyby(~HH_has_ITN, ~HH_head_education, DHS_SRS_wtdhh, svymean, na.rm=TRUE)
eduhh.SRS
```
```{r SRS weight2}
DHS_SRS_wtdchild<-svydesign(ids= ~1, weights=~pw,data=zam_child, nest=TRUE) #inclusion of ids= ~1 indicates no clustering (remember we are still assuming SRS at this point)

total.srswt.child<-svymean(~Slept_ITN, DHS_SRS_wtdchild)
total.srswt.child
```

``` {r SRS wtd prop2}


reschild.SRS <- svyby(~Slept_ITN, ~HH_residence, DHS_SRS_wtdchild, svymean, na.rm=TRUE) #Residence
reschild.SRS

seschild.SRS<- svyby(~Slept_ITN, ~HH_wealth_index, DHS_SRS_wtdchild, svymean, na.rm=TRUE)
seschild.SRS

educhild.SRS<- svyby(~Slept_ITN, ~Mother_education, DHS_SRS_wtdchild, svymean, na.rm=TRUE)
educhild.SRS

agechild.SRS <- svyby(~Slept_ITN, ~Child_age, DHS_SRS_wtdchild, svymean, na.rm=TRUE)
agechild.SRS
```

## Step 4: Calculate proportions and standard errors of select indicators assuming 2-Stage Cluster sampling with weights 

We now need to re-specify our survey design to reflect 2-Stage cluster sampling with weights. We will use the same syntax as when creating the SRS with weights survey design function, however will need to update our "ids" to correspond with our primary sampling unit, so that R Studio can see that our data is now clustered. 

```{r Cluster weights}

DHS_Clstr_wtdhh<-svydesign(ids= ~Primary_sampling_unit, weights=~pw,data=zam_hh, nest=TRUE)

total.clstrwt.hh<-svymean(~HH_has_ITN, DHS_Clstr_wtdhh)
total.clstrwt.hh
```

``` {r Cluster wtd prop}


Reshh.Clstr <- svyby(~HH_has_ITN, ~HH_residence, DHS_Clstr_wtdhh, svymean, na.rm=TRUE) #Residence
Reshh.Clstr

Eduhh.Clstr<- svyby(~HH_has_ITN, ~HH_head_education, DHS_Clstr_wtdhh, svymean, na.rm=TRUE)
Eduhh.Clstr

seshh.Clstr<- svyby(~HH_has_ITN, ~HH_wealth_index, DHS_Clstr_wtdhh, svymean, na.rm=TRUE)
seshh.Clstr
```

```{r Cluster weights2}

DHS_Clstr_wtdchild<-svydesign(ids= ~Primary_sampling_unit, weights=~pw,data=zam_child, nest=TRUE)

total.clstrwt.child<-svymean(~Slept_ITN, DHS_Clstr_wtdchild)
total.clstrwt.child
```

``` {r Cluster wtd prop2}


Reschild.Clstr <- svyby(~Slept_ITN, ~HH_residence, DHS_Clstr_wtdchild, svymean, na.rm=TRUE) 
Reschild.Clstr

Educhild.Clstr<- svyby(~Slept_ITN, ~Mother_education, DHS_Clstr_wtdchild, svymean, na.rm=TRUE)
Educhild.Clstr

seschild.Clstr<- svyby(~Slept_ITN, ~HH_wealth_index, DHS_Clstr_wtdchild, svymean, na.rm=TRUE)
seschild.Clstr

agechild.Clstr<- svyby(~Slept_ITN, ~Child_age, DHS_Clstr_wtdchild, svymean, na.rm=TRUE)
agechild.Clstr
```

## Step 5: Calculate proportions and standard errors of select indicators assuming 2-Stage Cluster sampling with weights and stratification

Here we need to create another function to implement/apply our survey design (2-Stage Cluster sampling with stratification) when calculating our proportions and standard errors.

```{r CWS}

DHS_Clstr_strathh <-svydesign(ids= ~Primary_sampling_unit, strata= ~Sample_stratum_number, weights=~pw,data=zam_hh, nest=TRUE)

total.clstrwtstrat.hh<-svymean(~HH_has_ITN, DHS_Clstr_strathh)
total.clstrwtstrat.hh

#note here that we have not specified 2-stage cluster sampling. As multistage sampling has no effect when fpc is not provided (for example with this data set), we do not need to specific 2-stage cluster sampling here when defining our "ids".
```


```{r CWS props}

Reshh.Strat <- svyby(~HH_has_ITN, ~HH_residence, DHS_Clstr_strathh, svymean, na.rm=TRUE) #Residence
Reshh.Strat

Eduhh.Strat <- svyby(~HH_has_ITN, ~HH_head_education, DHS_Clstr_strathh, svymean, na.rm=TRUE) 
Eduhh.Strat

Seshh.Strat <- svyby(~HH_has_ITN, ~HH_wealth_index, DHS_Clstr_strathh, svymean, na.rm=TRUE) 
Seshh.Strat

```
```{r CWS2}

DHS_Clstr_stratchild <-svydesign(ids= ~Primary_sampling_unit, strata= ~Sample_stratum_number, weights=~pw,data=zam_child, nest=TRUE)

DHS_Clstr_stratchildtotal<-svymean(~Slept_ITN, DHS_Clstr_stratchild)
DHS_Clstr_stratchildtotal


#note here that we have not specified 2-stage cluster sampling. As multistage sampling has no effect when fpc is not provided (for example with this data set), we do not need to specific 2-stage cluster sampling here when defining our "ids".
```


```{r CWS props2}

Reschild.Strat <- svyby(~Slept_ITN, ~HH_residence, DHS_Clstr_stratchild, svymean, na.rm=TRUE) #Residence
Reschild.Strat

Educhild.Strat <- svyby(~Slept_ITN, ~Mother_education, DHS_Clstr_stratchild, svymean, na.rm=TRUE) 
Educhild.Strat

Seschild.Strat <- svyby(~Slept_ITN, ~HH_wealth_index, DHS_Clstr_stratchild, svymean, na.rm=TRUE) 
Seschild.Strat

agechild.Strat <- svyby(~Slept_ITN, ~Child_age, DHS_Clstr_stratchild, svymean, na.rm=TRUE) 
agechild.Strat
```